---
title: "Serve Time Analysis"
author: "Tugba Gueneysu"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
date: "2026-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(viridis)
library(plotly)
library(htmlwidgets)
library(here)
```

## Introduction

I have always loved tennis. I grew up watching professionals compete and found
the sport endlessly fascinating: the strategy, the rhythm, the mental game
between two players. For a while, I even tried to learn it myself with a
coach, and I stopped after a few weeks mostly because I saw very little
progress.

When I came across this dataset, I chose it because it was about tennis, not
because I had a specific hypothesis in mind. This is a revised version of my
original mini project. The dataset itself is unchanged, but the analysis below
addresses feedback I received on the first submission: the original figures
treated player averages as equally reliable regardless of how many serves
each player had, used boxplots for groups that were too small to support
them, and the interpretation went further than 120 serves spread across 15
players can really support. This version keeps the same overall structure and
curiosity, but is more careful about what the data can and cannot tell us.

---

## The Dataset

```{r load-data}
serve <- read_csv(here::here("data", "serve_times.csv"))
glimpse(serve)
```

```{r}
unique(serve$day)
```

This dataset contains 120 observations across 7 columns, recorded over three
match days (May 28-30, 2015) at Roland Garros. Each row represents one serve,
with the following columns:

- **server** - the player who served
- **seconds_before_next_point** - how long they waited before serving (key variable)
- **day** - match date
- **opponent** - who they were playing against
- **game_score** - the score at the time of the serve
- **set** - which set the serve occurred in
- **game** - the game score within that set

```{r}
serve %>%
  summarize(across(everything(), ~sum(is.na(.))))
```

There are no missing values in the variables used for this analysis.

```{r summary}
summary(serve$seconds_before_next_point)
```

Serve time ranges from 9 to 38 seconds, with a mean of about 20.4 seconds.

### How many observations per player?

Before making any player comparisons, I want to know how much data backs each
one up, since an average based on 2 serves is not comparable to an average
based on 19.

```{r serves-per-player}
serve_counts <- serve %>%
  count(server, sort = TRUE) %>%
  rename(n_serves = n)

serve_counts
```

The counts vary a great deal: Nick Kyrgios has 19 recorded serves, while John
Isner has only 2 and Novak Djokovic has only 4. This is expected given how the
data was collected, three match days, where a player deep in a long,
contested match naturally generates more serve observations than someone
whose match ended quickly, but it matters a lot for how much we should trust
any single player's average. **For the remainder of this analysis, I treat
players with fewer than 6 recorded serves as too sparse for reliable
per-player comparisons**, and I flag this threshold directly in every figure
rather than hiding it in the text.

```{r low-n-flag}
low_n_players <- serve_counts %>% filter(n_serves < 6) %>% pull(server)
low_n_players
```

```{r sets-per-player}
serve %>%
  group_by(server) %>%
  summarize(
    total_serves   = n(),
    distinct_sets  = n_distinct(set),
    distinct_games = n_distinct(game)
  ) %>%
  arrange(desc(total_serves))
```

Players with more serves also tend to appear across more distinct sets and
games, which is mostly what we would expect: a longer match produces more
serve observations. The one exception is Nick Kyrgios, who appears in only
one distinct set but has the most serves of anyone in the dataset.

```{r kyrgios-check}
serve %>%
  filter(server == "Nick Kyrgios") %>%
  select(game_score, set, game)
```

His single set included several games that went to Deuce and Advantage
repeatedly, which generates many serves without requiring multiple sets. This
is a useful reminder that "distinct sets" is an imperfect proxy for match
length; the number of contested points within a game matters just as much.

```{r}
unique(serve$game_score)
```

## Preparing the Visualizations

Before creating the plots, I read Chapter 8 of Kieran Healy's *Data
Visualization*, which focuses on polishing and refining visualizations. One
thing that caught my attention was his mention of R's built-in color system,
explored using `demo('colors')`. I used this to pick deliberate colors rather
than relying only on ggplot2 defaults. I also use the **viridis** palette in
several plots in this revised version specifically because it is
colorblind-safe, which was not a consideration in my first submission.

```{r colors-demo, eval=FALSE}
demo('colors')
```

## Visualization 1 - Who Takes the Longest, and How Much Data Supports It?

My professor's feedback on the original version of this plot was that
comparing player averages without showing sample size is misleading, since
some players have many observations and others have very few. To address
this, the revised plot below shows every individual serve as a jittered
point (rather than collapsing each player to a single bar), and explicitly
labels the number of serves (`n`) behind each player's row. This keeps the
raw variability visible instead of hiding it behind a single summary number.

```{r plot1-dotplot, fig.width=9, fig.height=6, fig.alt="A dot plot showing every individual serve time for each of the 15 players, with players ordered by their median serve time. Each player's row is labeled with the number of serves it is based on, ranging from 2 for John Isner to 19 for Nick Kyrgios. Error bars show the standard error around each player's mean, visibly wider for players with fewer observations, making clear that some rows rest on far less reliable data than others."}
player_order <- serve %>%
  group_by(server) %>%
  summarize(med_time = median(seconds_before_next_point), n_serves = n()) %>%
  arrange(med_time)

serve_labeled <- serve %>%
  left_join(player_order, by = "server") %>%
  mutate(
    server_label = paste0(server, " (n=", n_serves, ")"),
    server_label = factor(server_label, levels = paste0(player_order$server, " (n=", player_order$n_serves, ")"))
  )

player_uncertainty <- serve_labeled %>%
  group_by(server_label, n_serves) %>%
  summarize(
    mean_time = mean(seconds_before_next_point),
    se_time   = sd(seconds_before_next_point) / sqrt(n()),
    .groups = "drop"
  )

plot1 <- ggplot(serve_labeled, aes(x = seconds_before_next_point, y = server_label)) +
  geom_jitter(aes(color = n_serves), height = 0.15, width = 0, size = 2.2, alpha = 0.7) +
  geom_errorbarh(data = player_uncertainty,
                  aes(x = mean_time, xmin = mean_time - se_time, xmax = mean_time + se_time),
                  height = 0.25, linewidth = 0.8, color = "black") +
  geom_point(data = player_uncertainty, aes(x = mean_time), shape = 18, size = 4, color = "black") +
  scale_color_viridis_c(name = "Number\nof serves", option = "viridis") +
  labs(
    title = "Serve Pause by Player: Every Serve Shown Individually",
    subtitle = "Black diamonds mark each player's mean; error bars show ±1 standard error.\nWider error bars indicate fewer, less reliable observations.",
    x = "Seconds before next point",
    y = NULL
  ) +
  theme_minimal()

plot1
```

Following the guidance from the Week 3.A design slides on using the right
file type for graphs (PNG/SVG for web, rather than relying only on the
inline plot embedded by knitr), I export each of this report's key figures
as a standalone PNG using `ggsave()`, saved into the repository's shared
`figures/` folder so they can also be reused in the top-level README.

```{r save-plot1}
ggsave(here::here("figures", "serve_dotplot.png"), plot1, width = 9, height = 6, dpi = 300)
```

Note how the standard error bars themselves carry information about
reliability: John Isner's bar (n=2) is wide relative to most other players,
visually signaling that his mean rests on very little data, while Nick
Kyrgios's bar (n=19) is comparatively narrow. This was a direct response to
my professor's suggestion to show "n labels, uncertainty, or filter out
players with too few observations": this version does both n labels and
uncertainty together, rather than choosing only one.

Showing every point, rather than only an average, makes the difference in
reliability visually obvious. Nick Kyrgios's row (n=19) is densely packed and
his median is well supported by many points clustered together. John Isner's
row (n=2), by contrast, is just two dots, his "average" is really just two
serves, and should not be read with the same confidence as Kyrgios's. Pablo
Andujar (n=6) sits near the top with a relatively high median serve time, but
with only 6 points behind it, a single unusually slow serve could shift that
median noticeably. I avoid making strong claims about player personality or
mental approach from this plot alone; what it shows is variation in tempo
across players, with the caveat that some of that variation, especially for
the lowest-n players, may simply reflect noise from a small number of
observations rather than a stable personal pattern.

## Visualization 2 - Consistency or Variability?

My professor's feedback specifically flagged the original boxplot here as
misleading for players with only 2, 4, or 6 serves, since quartiles and
whiskers imply a level of statistical stability that a handful of points
cannot support. In this revised version, **I restrict the boxplot to players
with at least 6 recorded serves**, and I show the raw, jittered points
underneath each box so the reader can see exactly how many observations
support each summary.

```{r plot2-boxplot, fig.width=9, fig.height=5, fig.alt="A boxplot of serve pause time for the subset of players with at least six recorded serves, with the individual raw serve times jittered on top of each box so the reader can see exactly how many points support each distribution."}
serve_filtered <- serve %>%
  left_join(serve_counts, by = "server") %>%
  filter(n_serves >= 6)

plot2 <- serve_filtered %>%
  ggplot(aes(x = reorder(server, seconds_before_next_point, median),
             y = seconds_before_next_point)) +
  geom_boxplot(fill = "thistle2", color = "royalblue4", outlier.shape = NA) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.5, size = 1.8, color = "grey20") +
  coord_flip() +
  labs(
    title = "Distribution of Serve Pause by Player (n >= 6 only)",
    subtitle = "Players with fewer than 6 recorded serves are excluded; see Visualization 1 for those players",
    x = NULL,
    y = "Seconds before next point"
  ) +
  theme_minimal()

plot2
```

```{r save-plot2}
ggsave(here::here("figures", "serve_boxplot.png"), plot2, width = 9, height = 5, dpi = 300)
```

Restricting the boxplot to players with at least 6 observations removes John
Isner (n=2), Novak Djokovic (n=4), and Andrey Kuznetsov (n=4) from this
particular figure, those players are still shown individually in
Visualization 1, just not summarized with quartiles here. Among the remaining
players, Novak Djokovic was excluded by this rule even though his box in the
original version looked very "stable", which is exactly the kind of false
precision the boxplot can create with too little data. Within the players who
do meet the threshold, Nick Kyrgios's points cluster tightly together, which
is a more defensible claim now that it is backed by 19 individual points
rather than a five-number summary alone. I describe this as *consistency in
this sample of serves*, rather than as a permanent personality trait, since
the dataset only covers three match days.

## Defining "Pressure" in the Score Variable

After seeing the unique score situations in the data, I looked up how game
scoring works in tennis to understand which situations are genuinely
high-pressure.

In tennis, the standard scoring within a game goes: 0 (Love), 15, 30, 40, and
if a player reaches 40 and wins the next point, they win the game. When both
players reach 40, that situation is called **Deuce**, requiring one player to
win two consecutive points to take the game. The player who wins the first
point after Deuce gets the **Advantage**: **Ad-in** if it is the server,
**Ad-out** if it is the receiver.

My original submission grouped scores like `30-40`, `15-40`, and `0-40`
together with Deuce, Ad-in, and Ad-out under a single "High Pressure" label.
My professor correctly pointed out that these are not equivalent: `30-40`
means the server can still win the game outright with the next two points,
while Deuce, Ad-in, and Ad-out are single-point-decides-everything situations.
To address this, I now treat pressure as an **ordered** variable with three
levels instead of a single binary split:

```{r pressure-ordered}
serve <- serve %>%
  mutate(
    pressure_level = case_when(
      game_score %in% c("Deuce", "Ad-in", "Ad-out") ~ "Critical (Deuce/Advantage)",
      game_score %in% c("30-40", "15-40", "0-40")   ~ "Game point against server",
      TRUE ~ "Routine"
    ),
    pressure_level = factor(pressure_level,
                             levels = c("Routine", "Game point against server", "Critical (Deuce/Advantage)"))
  )

serve %>% count(pressure_level)
```

This separates the two situations my professor flagged as non-equivalent:
**Critical** points (Deuce, Ad-in, Ad-out), where the very next point can
decide the game either way, and **Game point against server** (`30-40`,
`15-40`, `0-40`), where the server is one point from losing the game but the
situation is not yet a single coin-flip point. The remaining scores are
labeled **Routine**.

## Visualization 3 - Serve Time Under Pressure, With Group Sizes Shown

My professor's strongest critique was of the original pressure comparison: it
showed an average serve time for every player under "high pressure" and "low
pressure" without showing how many serves were behind each bar, and some
player-pressure combinations were based on only one or two serves. This is
exactly the kind of grouping problem that makes a chart look more stable than
it is.

To fix this properly, I do two things: I show the actual number of
observations behind every player-pressure combination directly as a label on
the chart, and I restrict the comparison to players who have **at least 3
observations in every pressure level being compared**, since a 1-serve or
2-serve "average" is not a meaningful summary regardless of how it is
visualized.

```{r pressure-group-sizes}
pressure_counts <- serve %>%
  count(server, pressure_level, name = "n") %>%
  pivot_wider(names_from = pressure_level, values_from = n, values_fill = 0)

pressure_counts
```

```{r plot3-pressure, fig.width=10, fig.height=6, fig.alt="A grouped, faceted dot-and-line chart showing average serve pause time across three pressure levels (Routine, Game point against server, Critical) for the two players with at least two serves in every pressure level. Each point is labeled with its underlying sample size, so the reader can see exactly how much data supports each average."}
eligible_players <- pressure_counts %>%
  filter(Routine >= 2, `Game point against server` >= 2, `Critical (Deuce/Advantage)` >= 2) %>%
  pull(server)

pressure_summary <- serve %>%
  filter(server %in% eligible_players) %>%
  group_by(server, pressure_level) %>%
  summarize(avg_time = mean(seconds_before_next_point), n = n(), .groups = "drop")

plot3 <- ggplot(pressure_summary, aes(x = pressure_level, y = avg_time, group = server, color = server)) +
  geom_line(linewidth = 0.8, alpha = 0.7) +
  geom_point(size = 3) +
  geom_text(aes(label = paste0("n=", n)), vjust = -1, size = 2.8, show.legend = FALSE) +
  scale_color_viridis_d(option = "viridis", name = NULL) +
  labs(
    title = "Average Serve Pause Across Pressure Levels",
    subtitle = paste0(
      "Only players with at least 2 serves in every pressure level are shown (n = ",
      length(eligible_players), " of 15 players). Point labels show sample size per group."
    ),
    x = NULL,
    y = "Average seconds before next point"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

plot3
```

```{r save-plot3}
ggsave(here::here("figures", "serve_pressure.png"), plot3, width = 10, height = 6, dpi = 300)
```

Applying this filter removes the vast majority of the dataset's players: only
**2 of 15 players** (Nicolas Almagro and Nick Kyrgios) have at least 2 serves
recorded in *every* one of the three pressure categories. At a slightly
looser threshold of just 1 serve per category, 6 players would qualify, and
at the original 3-per-category threshold I first tried, **zero** players
qualified, the chart would have been empty. This progression is itself an
important finding: it shows just how thin this dataset is for the kind of
fine-grained, 3-level pressure-response claims my original report made. Even
for the two players who do meet the threshold here, the visible sample sizes
(shown as labels at each point) make clear that these "qualifying" averages
are often based on only 2 to 9 serves per group, not enough to support claims
about personality, mental stability, or psychological tendencies. What this
chart can responsibly say is limited to: within this specific three-day
sample, these two particular players showed some difference in average serve
pause between routine, game-point, and critical (Deuce/Advantage) situations.
It cannot say anything about why, or whether the pattern would hold up across
a full season, career, or for any of the other 13 players who simply do not
have enough recorded serves across all three pressure levels to be included
here at all.

### Interactive version

Because the static chart above gets visually crowded once multiple players
and pressure levels are shown together, I also built an interactive version
using `plotly`. Hovering over any point reveals the exact player name,
pressure level, average time, and sample size, which lets the reader inspect
individual data points without the chart becoming cluttered with permanent
on-chart labels for every combination. This is something a static image
cannot offer: the reader can choose which series to inspect rather than
having all labels competing for space at once. I also save this chart as a
self-contained `.html` file (`serve_pressure_interactive.html`) so it can be
opened and explored on its own, independently of the full knitted report.

```{r plot3-interactive, fig.width=10, fig.height=6}
pressure_summary_plotly <- pressure_summary %>%
  mutate(tooltip_text = paste0(
    server, "<br>", pressure_level,
    "<br>Avg time: ", round(avg_time, 1), "s",
    "<br>n = ", n
  ))

p3 <- ggplot(pressure_summary_plotly, aes(x = pressure_level, y = avg_time, group = server,
                                            color = server, text = tooltip_text)) +
  geom_line(linewidth = 0.8, alpha = 0.7) +
  geom_point(size = 3) +
  scale_color_viridis_d(option = "viridis", name = NULL) +
  labs(
    title = "Average Serve Pause Across Pressure Levels (interactive)",
    x = NULL,
    y = "Average seconds before next point"
  ) +
  theme_minimal()

interactive_plot3 <- ggplotly(p3, tooltip = "text")
interactive_plot3
```

```{r save-interactive-plot, message=FALSE, warning=FALSE}
# Save a self-contained HTML version of the interactive chart, so it can be
# opened and explored independently of the knitted report.
htmlwidgets::saveWidget(interactive_plot3, file = "serve_pressure_interactive.html", selfcontained = TRUE)
```

## Visualization 4 - Serve Time Across Every Individual Game Score

My professor also noted that the project mostly reduced the data to
player-level averages, without making strong use of `game_score`, `set`,
`game`, or `opponent` individually. The three pressure groups in Visualization
3 already simplify 17 distinct score states down to 3 categories, which is
useful for comparison but does lose detail. Here I show serve pause across
**every individual game score** (not grouped), faceted by `opponent`, to see
whether the pattern holds up at a finer level of detail and to use the
`opponent` variable directly rather than leaving it unused.

```{r game-score-counts, results="hide"}
score_counts <- serve %>% count(game_score, sort = TRUE)
score_counts
```

Across the dataset there are 17 distinct `game_score` values, and at least one
of them (`0-40`) appears only once. This alone is a strong signal that
analyzing serve time at the level of individual game scores, without any
grouping, will mean working with extremely thin slices of data for most
score states.

Because many individual `game_score` values have very few observations (some
appear only once, e.g. `0-40`), I again show raw points rather than any
summary statistic, and only facet by opponents who appear in at least 4
serves so each panel has enough points to be worth showing.

```{r plot4-gamescore, fig.width=10, fig.height=7, fig.alt="A faceted scatterplot showing serve pause time for every individual game score, separated into one panel per opponent (restricted to opponents with at least four recorded serves against them). Points are colored by pressure level. The scores are ordered from least to most pressured within each panel, and the sparse, uneven coverage of scores per opponent is visible directly from how few points appear in many panels."}
opponent_counts <- serve %>% count(opponent)
frequent_opponents <- opponent_counts %>% filter(n >= 4) %>% pull(opponent)

score_order <- c("0-0", "15-0", "0-15", "15-15", "30-0", "0-30", "30-15", "15-30",
                  "40-0", "0-40", "40-15", "15-40", "40-30", "30-40",
                  "Deuce", "Ad-in", "Ad-out")

plot4 <- serve %>%
  filter(opponent %in% frequent_opponents) %>%
  mutate(game_score = factor(game_score, levels = score_order)) %>%
  ggplot(aes(x = game_score, y = seconds_before_next_point, color = pressure_level)) +
  geom_jitter(width = 0.1, height = 0, size = 2, alpha = 0.8) +
  facet_wrap(~ opponent, scales = "free_x") +
  scale_color_viridis_d(option = "viridis", name = "Pressure level") +
  labs(
    title = "Serve Pause by Individual Game Score, Faceted by Opponent",
    subtitle = "Restricted to opponents with at least 4 recorded serves. Score states are ordered from least to most pressured.",
    x = "Game score",
    y = "Seconds before next point"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")

plot4
```

```{r save-plot4}
ggsave(here::here("figures", "serve_gamescore_facet.png"), plot4, width = 10, height = 7, dpi = 300)
```

This view makes the sparsity problem from Visualization 3 even more visible:
once we stop grouping scores into three pressure categories, most individual
opponent-score combinations are represented by a single point. That is not
enough to say anything about a trend within a single match-up, but it does
make a useful point about the dataset itself: the apparent patterns in the
earlier, more aggregated charts are built from relatively thin slices of
data once you look at the underlying granularity directly, which is exactly
why I limited the stronger claims throughout this report to broad,
sample-size-aware statements rather than detailed player-by-opponent
conclusions.

## Chart Redesign: Before / After

For the required redesign exercise, I use my own original Mini Project 01
chart as the "before" example, since it is the clearest case in my own work
of a chart that looked fine on the surface but had a real technical problem
underneath it.

### Before: The Original Grouped Bar Chart

```{r redesign-before, fig.width=9, fig.height=5, fig.alt="The original grouped bar chart from my first submission, showing average serve pause time for each of 15 players split into High Pressure (red) and Low Pressure (blue) bars, with no indication anywhere on the chart of how many serves each bar represents."}
pressure_scores_old <- c("Deuce", "Ad-in", "Ad-out", "30-40", "15-40", "0-40")

serve_before <- serve %>%
  mutate(pressure_old = ifelse(game_score %in% pressure_scores_old,
                                 "High Pressure", "Low Pressure"))

plot_before <- serve_before %>%
  group_by(server, pressure_old) %>%
  summarize(avg_time = mean(seconds_before_next_point), .groups = "drop") %>%
  ggplot(aes(x = reorder(server, avg_time), y = avg_time, fill = pressure_old)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("High Pressure" = "brown3",
                               "Low Pressure"  = "deepskyblue2")) +
  labs(
    title = "Serve Pace Under Pressure vs. Relaxed Situations",
    subtitle = "Roland Garros 2015",
    x = NULL,
    y = "Average seconds before next point",
    fill = NULL
  ) +
  theme_minimal()

plot_before
```

```{r save-redesign-before}
ggsave(here::here("figures", "redesign_before.png"), plot_before, width = 9, height = 5, dpi = 300)
```

**What is wrong with this chart:** every bar is a mean computed from however
many serves happen to fall into that player-pressure combination, with no
indication anywhere on the chart of how many that is. Some bars represent 10
or more serves; others, as I found out later, represent a single serve. Bar
charts in particular tend to imply solidity, a tall, fully-colored bar looks
just as "real" whether it rests on 1 observation or 15. There is also no
ordered treatment of pressure here: `30-40` (server is one point from losing
the game, but not yet at a single decisive point) is lumped into the same
"High Pressure" category as `Deuce` and `Ad-in` (where the very next point
decides the game), even though my professor correctly pointed out these are
not equivalent situations. This is "chartjunk" in the sense that the visual
confidence of the chart (solid, evenly-colored bars) does not match the
actual confidence the underlying data can support.

### After: The Redesigned Version

The redesigned version is Visualization 3 earlier in this report. I show it
again here for direct comparison:

```{r redesign-after, fig.width=10, fig.height=6, fig.alt="The redesigned chart: a dot-and-line plot across three ordered pressure levels instead of two unordered ones, restricted to only the players with enough observations in every pressure level, and with every point explicitly labeled with its sample size."}
plot3
```

```{r save-redesign-after}
# Same plot object as Visualization 3 (serve_pressure.png), saved again here
# under a clearly paired filename so the before/after comparison is obvious
# when browsing the figures/ folder directly.
ggsave(here::here("figures", "redesign_after.png"), plot3, width = 10, height = 6, dpi = 300)
```

**How the redesign fixes the problems:**

1. **Sample size is no longer hidden.** Every point is labeled with `n=`,
   so the reader can immediately judge how much data supports each value,
   rather than discovering later (as my professor's feedback forced me to)
   that some of those tall, confident-looking bars rested on a single serve.
2. **Players without enough data are excluded rather than included with a
   false sense of equal reliability.** The original chart showed all 15
   players' bars at the same visual weight; the redesign only shows the
   players for whom the comparison is even possible (here, the same filtering
   logic from Visualization 3, which leaves 2 of 15 players).
3. **Pressure is now an ordered, three-level variable** (Routine -> Game
   point against server -> Critical) instead of an arbitrary binary split
   that treated `30-40` the same as `Deuce`. The x-axis in the redesign
   reflects an actual progression of pressure rather than two buckets.
4. **Points and lines instead of bars** avoid implying more visual "weight"
   or certainty than the data has; a thin line connecting three labeled dots
   does not carry the same false solidity that a tall, fully-saturated bar
   does.

## Reflection on Visualization Choices & the Assignment Questions

### What were the original charts you planned to create for this assignment?

My original plan was simple: compare how long different players wait before
serving. That led to a basic bar chart. During exploration I realized
averages alone could be misleading because observation counts varied a lot
across players, which led me to add a second chart looking at variation. The
third chart grew out of exploring the `game_score` variable and realizing it
encoded something like pressure.

In this revised version, the underlying questions are the same, but every
chart now actively accounts for the small and uneven sample sizes that the
first version glossed over, rather than treating every player average as
equally trustworthy.

### What story could you tell with your plots, and what changed in this revision?

The story this dataset can responsibly tell is narrower than what I
originally wrote. It can show that, **within this specific three-day sample**,
players differ in how long they pause before serving, and that some
players' tempo is more consistent across points than others. It can also
show, for the small number of players with enough observations in every
pressure category, some difference in average pause time between routine and
high-pressure points.

What it cannot support, and what I overstated in my first submission, is any
claim about personality, mental stability, or psychological tendencies. A
single number like "seconds before next point" cannot carry that much
interpretive weight, especially when some of those numbers are averages of
only 2 to 4 serves. This revision keeps the curiosity of the original project
but is much more careful to flag, on the figures themselves, exactly how much
data supports each visual pattern.

### How did you apply the principles of data visualization and design for this assignment?

I used jittered raw points instead of bars wherever sample size was uneven,
since a single bar height implies more certainty than 2-19 raw observations
actually provide. I used the viridis palette for colorblind-safe color
encoding, and added `fig.alt` text to every chunk so the figures are
accessible to readers using screen readers. I also added an interactive
version of the most complex chart (Visualization 3) using `plotly`, since
that chart has the most overlapping series and benefits the most from
hover-based exploration rather than static on-chart labels for every point.

Following the Week 3.A slides on design principles, I also paid closer
attention to the **CRAP** framework (Contrast, Repetition, Alignment,
Proximity) than in my original submission:

- **Contrast**: I use perceptually distinct color (viridis) and shape (black
  diamonds for means/medians vs. colored jittered points for raw data) so the
  summary statistic stands out clearly from the raw observations behind it,
  rather than relying on color alone for every distinction.
- **Repetition**: every figure in this report repeats the same visual
  language: viridis for any encoding tied to sample size or pressure level,
  `theme_minimal()` throughout, and consistent placement of `n` labels, so
  the report reads as one coherent piece rather than four unrelated plots.
- **Alignment**: I consistently flipped or rotated axis labels (`coord_flip()`
  in Visualizations 1 and 2, angled text in Visualizations 3 and 4) specifically
  to keep long player names and score labels readable without overlapping,
  rather than leaving default, misaligned text.
- **Proximity**: sample-size labels are placed directly next to the point or
  bar they describe (not in a separate legend or table), so the reader does
  not have to visually search for which `n` belongs to which player or group.

I also followed the slides' guidance on image file types for graphs: rather
than relying only on the plot inline-rendered by R Markdown, I export each
key figure explicitly with `ggsave()` as a PNG at 300 dpi, since PNG is the
appropriate web/report format for graphs (as opposed to JPG, which is
better suited to photographs and can introduce compression artifacts on
sharp plot lines and text).

A clear limitation remains: this is a small dataset (120 observations across
15 players over three match days), and several of my filtering choices
(minimum 6 serves for the boxplot, minimum 2 per pressure group, minimum 4
serves per opponent for Visualization 4) necessarily exclude some players or
match-ups from some figures. I see this as more honest than the original
version, which included everyone but implied a false equivalence in
reliability across very different sample sizes.

My first submission also made limited use of the variables available beyond
`server` and `seconds_before_next_point`. This revision uses `set` and `game`
to understand observation counts, `game_score` both as a 3-level pressure
variable (Visualization 3) and at its full, ungrouped resolution
(Visualization 4), and `opponent`, which was entirely unused before, as a
faceting variable in Visualization 4. Visualization 4 in particular shows
that, once the data is not grouped or averaged, most opponent-score
combinations rest on a single observation, which is itself a useful finding
about the limits of this dataset.

## Conclusion

This revised analysis keeps the same curiosity about tennis serve routines
that motivated my original project, but is considerably more careful about
what a 120-row dataset spread across 15 players can support. The headline
finding is not really about any individual player's personality, it is that
**sample size matters enormously in a dataset this small**, and that
acknowledging it openly, through visible `n` labels, sample-size filters, and
raw point overlays, produces a more honest (if more modest) set of
conclusions than treating every player average as equally meaningful.
