---
title: "Data Visualization Project 03 – Final"
author: "Tugba Gueneysu"
output:
  html_document:
    keep_md: true
    toc: true
    toc_float: true
date: "2026-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(ggridges)
library(viridis)
library(plotly)
library(htmlwidgets)
library(here)
```

In this exercise I explore methods to create different types of data
visualizations, including plotting text data and exploring the distributions
of continuous variables.

## PART 1: Density Plots

Using the dataset obtained from FSU's [Florida Climate Center](https://climatecenter.fsu.edu/climate-data-access-tools/downloadable-data),
for a station at Tampa International Airport (TPA) for 2022, I attempt to
recreate the charts shown in the assignment template, which were originally
generated using data from 2016.

```{r load-weather}
weather_tpa <- read_csv(here::here("data", "tpa_weather_2022.csv"))

# random sample
sample_n(weather_tpa, 4)
```

Before plotting, I convert `month` into a proper factor with month names
(using `lubridate`), and check for the missing-data placeholder codes
mentioned in the assignment (`-99.9` for temperatures, `-99.99` for
precipitation).

```{r clean-weather}
weather_tpa <- weather_tpa %>%
  mutate(
    date = make_date(year, month, day),
    month_name = factor(month.name[month], levels = month.name)
  )

weather_tpa %>%
  summarise(
    n_missing_temp = sum(max_temp == -99.9 | min_temp == -99.9, na.rm = TRUE),
    n_missing_precip = sum(precipitation == -99.99, na.rm = TRUE)
  )
```

The 2022 station data for TPA does not contain any of these missing-data
placeholder codes, so no rows needed to be filtered out before plotting.

### (a) Faceted histogram of maximum temperatures by month

```{r part1a, fig.alt="Twelve small histograms, one per month, showing the distribution of daily maximum temperatures at Tampa International Airport in 2022. Cooler months on the left show wider, lower, more spread-out distributions; summer months show tighter distributions concentrated in the high 80s to low 90s degrees Fahrenheit."}
plot1a <- ggplot(weather_tpa, aes(x = max_temp, fill = month_name)) +
  geom_histogram(binwidth = 3, color = "white", linewidth = 0.2) +
  scale_fill_viridis_d(option = "viridis", guide = "none") +
  facet_wrap(~ month_name, nrow = 3) +
  labs(
    title = "Monthly Maximum Temperature Distributions (TPA, 2022)",
    x = "Maximum temperatures",
    y = "Number of Days"
  ) +
  theme_minimal() +
  theme(strip.background = element_rect(fill = "grey85", color = NA))

plot1a
```

```{r save-plot1a}
ggsave(here::here("figures", "tpa_2022_max_temps_facet.png"), plot1a, width = 9, height = 6, dpi = 300)
```

This matches the structure of the target chart: warmer, summer months (June
through September) show narrow, tall histograms concentrated near 90°F,
while winter months (December through February) are flatter and more spread
out across the 60-80°F range.

### (b) Density plot of maximum temperatures (overall)

```{r part1b, fig.alt="A single density curve of daily maximum temperatures at Tampa International Airport in 2022, showing a bumpy, multi-modal shape with the largest peak near 90 degrees Fahrenheit and a smaller, broader bump in the 60s and 70s."}
plot1b <- ggplot(weather_tpa, aes(x = max_temp)) +
  geom_density(kernel = "rectangular", bw = 0.5, fill = "grey40", color = "black") +
  labs(
    title = "Density of Maximum Temperatures (TPA, 2022)",
    x = "Maximum temperature",
    y = "density"
  ) +
  theme_minimal()

plot1b
```

```{r save-plot1b}
ggsave(here::here("figures", "tpa_2022_max_temps_density.png"), plot1b, width = 9, height = 6, dpi = 300)
```

Using `bw = 0.5` with a rectangular kernel produces a visibly "bumpy",
high-variance density curve rather than a smooth one, since the bandwidth is
very narrow. This shows the same overall shape as the histogram in (a): one
broad, low bump in the 60s-70s range and one tall, sharp peak around 90°F.

### (c) Faceted density plots by month

```{r part1c, fig.alt="Twelve small density curves, one per month, showing the distribution of daily maximum temperatures at Tampa International Airport in 2022, colored from purple in January to yellow in December using the viridis palette."}
plot1c <- ggplot(weather_tpa, aes(x = max_temp, y = after_stat(density), fill = month_name)) +
  geom_density(color = "black") +
  scale_fill_viridis_d(option = "viridis", guide = "none") +
  facet_wrap(~ month_name, nrow = 3) +
  labs(
    title = "Density plots for each month in 2022",
    x = "Maximum temperatures",
    y = NULL
  ) +
  theme_minimal() +
  theme(strip.background = element_rect(fill = "grey85", color = NA))

plot1c
```

```{r save-plot1c}
ggsave(here::here("figures", "tpa_2022_max_temps_density_facet.png"), plot1c, width = 9, height = 6, dpi = 300)
```

With default `geom_density()` settings (smoother bandwidth than part (b)),
each month's distribution becomes easier to read individually. The
progression from winter to summer is visible as the peak shifts to the right
and becomes narrower and taller, consistent with both the histogram in (a)
and the single density curve in (b).

### (d) Ridgeline plot of maximum temperatures by month

```{r part1d, fig.width=8, fig.height=6, fig.alt="A ridgeline plot showing the distribution of daily maximum temperatures for each month of 2022 in Tampa, stacked from January at the bottom to December at the top. Each ridge is colored on a plasma gradient from dark purple at cooler temperatures to bright yellow at warmer temperatures, with a vertical line marking the median temperature for that month."}
plot1d <- ggplot(weather_tpa, aes(x = max_temp, y = month_name, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    quantile_lines = TRUE,
    quantiles = 2
  ) +
  scale_fill_viridis_c(option = "plasma", name = "Max temp (°F)") +
  labs(
    title = "Temperature Distribution by Month (TPA, 2022)",
    x = "Maximum temperature (in Fahrenheit degrees)",
    y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "right")

plot1d
```

```{r save-plot1d}
ggsave(here::here("figures", "tpa_2022_max_temps_ridges_plasma.png"), plot1d, width = 8, height = 6, dpi = 300)
```

The `quantile_lines = TRUE` and `quantiles = 2` settings add a vertical line
at each month's median, which makes the rightward shift from winter to
summer easier to track precisely across the 12 stacked ridges, rather than
just eyeballing where each curve peaks.

### (e) Visualizing precipitation

For this plot of my own choice, I look at total monthly precipitation
alongside the number of rainy days per month, since a single density plot of
daily precipitation would be dominated by the many zero-precipitation days
and would not tell an interesting story on its own.

```{r part1e, fig.alt="A bar chart showing total precipitation in inches for each month of 2022 at Tampa International Airport, colored using a colorblind-safe viridis palette, with the wettest months in late summer standing out clearly from the drier winter and spring months."}
monthly_precip <- weather_tpa %>%
  group_by(month_name) %>%
  summarise(
    total_precip = sum(precipitation, na.rm = TRUE),
    rainy_days = sum(precipitation > 0, na.rm = TRUE)
  )

plot1e <- ggplot(monthly_precip, aes(x = month_name, y = total_precip, fill = total_precip)) +
  geom_col() +
  scale_fill_viridis_c(option = "viridis", guide = "none") +
  labs(
    title = "Total Monthly Precipitation in Tampa, 2022",
    x = NULL,
    y = "Total precipitation (inches)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot1e
```

```{r save-plot1e}
ggsave(here::here("figures", "tpa_2022_precipitation.png"), plot1e, width = 9, height = 6, dpi = 300)
```

The chart shows that 2022 precipitation in Tampa was concentrated in a
handful of months (especially mid-summer), rather than spread evenly across
the year, consistent with Florida's wet-season/dry-season pattern.

## PART 2: Data on Concrete Strength (Option B)

Concrete is the most important material in **civil engineering**. The
concrete compressive strength is a highly nonlinear function of *age* and
*ingredients*. The dataset used here is from the [UCI Machine Learning
Repository](https://archive.ics.uci.edu/ml/index.php), and it contains 1030
observations with 9 attributes (8 quantitative input variables and 1
quantitative output variable). A data dictionary is included below:

Variable                      |    Notes                
------------------------------|-------------------------------------------
Cement                        | kg in a $m^3$ mixture             
Blast Furnace Slag            | kg in a $m^3$ mixture  
Fly Ash                       | kg in a $m^3$ mixture             
Water                         | kg in a $m^3$ mixture              
Superplasticizer              | kg in a $m^3$ mixture
Coarse Aggregate              | kg in a $m^3$ mixture
Fine Aggregate                | kg in a $m^3$ mixture      
Age                           | in days                                             
Concrete compressive strength | MPa, megapascals

```{r load-concrete}
concrete <- read_csv(here::here("data", "concrete.csv"), col_types = cols())
```

I create a new attribute for visualization purposes, `strength_range`, which
bins compressive strength into quintiles:

```{r strength-range}
new_concrete <- concrete %>%
  mutate(strength_range = cut(Concrete_compressive_strength,
                               breaks = quantile(Concrete_compressive_strength,
                                                  probs = seq(0, 1, 0.2)),
                               include.lowest = TRUE))
```

### 1. Distribution of two continuous variables

```{r part2-1-cement, fig.alt="A histogram showing the distribution of cement content across the 1030 concrete samples, right-skewed with most samples between roughly 150 and 350 kilograms per cubic meter and a long tail toward higher values."}
plot2_1a <- ggplot(new_concrete, aes(x = Cement)) +
  geom_histogram(binwidth = 20, fill = viridis(1), color = "white") +
  labs(
    title = "Distribution of Cement Content",
    x = "Cement (kg in a m³ mixture)",
    y = "Number of samples"
  ) +
  theme_minimal()

plot2_1a
```

```{r part2-1-strength, fig.alt="A histogram showing the distribution of concrete compressive strength across the 1030 samples, roughly bell-shaped but slightly right-skewed, ranging from near zero to over 80 megapascals."}
plot2_1b <- ggplot(new_concrete, aes(x = Concrete_compressive_strength)) +
  geom_histogram(binwidth = 5, fill = viridis(1, option = "magma"), color = "white") +
  labs(
    title = "Distribution of Concrete Compressive Strength",
    x = "Compressive strength (MPa)",
    y = "Number of samples"
  ) +
  theme_minimal()

plot2_1b
```

```{r save-plot2-1}
ggsave(here::here("figures", "concrete_cement_distribution.png"), plot2_1a, width = 7, height = 5, dpi = 300)
ggsave(here::here("figures", "concrete_strength_distribution.png"), plot2_1b, width = 7, height = 5, dpi = 300)
```

**Do the ranges make sense?** Cement content ranges from about 100 to over
500 kg/m³, and the distribution is right-skewed: most mixtures use a
moderate amount of cement, while a smaller number of mixtures use
considerably more. Compressive strength ranges from close to 0 MPa to over
80 MPa, with most samples falling between roughly 20 and 50 MPa. Both ranges
make sense for concrete: very low strength values correspond to young or
poorly cured samples, and the maximum of around 80 MPa is consistent with
high-strength concrete mixtures, which can realistically reach that range in
practice.

### 2. Strength by age

```{r part2-2, fig.alt="A boxplot showing concrete compressive strength at each distinct age value measured in days, colored by quintile strength range, showing that strength generally increases with age up to around 90 days and then levels off or even decreases slightly at very long ages such as 270 or 365 days."}
plot2_2 <- ggplot(new_concrete, aes(x = factor(Age), y = Concrete_compressive_strength, fill = strength_range)) +
  geom_boxplot() +
  scale_fill_viridis_d(option = "viridis", name = "Strength Range") +
  labs(
    title = "Concrete Compressive Strength by Age",
    x = "Age (in days)",
    y = "Compressive Strength (in MPa)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot2_2
```

```{r save-plot2-2}
ggsave(here::here("figures", "concrete_strength_by_age.png"), plot2_2, width = 10, height = 6, dpi = 300)
```

**Comment on results:** compressive strength increases sharply over the
first weeks of curing, which matches the well-known chemistry of concrete:
most of the strength gain happens in the first 28 days as the cement
hydrates, and gains continue more slowly afterward. Strength levels off by
around 90 days, and the very long-age samples (270, 365 days) do not show
much further increase, and in some cases show high variability instead. This
makes sense, since very-long-age samples in this dataset are fewer in number
and may include different mixture compositions than the more commonly-tested
ages like 28 or 56 days.

### 3. Strength versus cement, age, and water

```{r part2-3-static, fig.alt="A scatterplot of compressive strength against cement content, where each point is sized by age in days and colored by water content using a colorblind-safe viridis scale; higher cement content is generally associated with higher strength, and points with less water tend to cluster toward higher strength values."}
plot2_3 <- ggplot(new_concrete, aes(x = Cement, y = Concrete_compressive_strength, size = Age, color = Water)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(option = "viridis") +
  labs(
    title = "Exploring Strength versus (Cement, Age, and Water)",
    x = "Cement",
    y = "Strength",
    size = "Age",
    color = "Water",
    caption = "Age is measured in days"
  ) +
  theme_minimal()

plot2_3
```

```{r save-plot2-3}
ggsave(here::here("figures", "concrete_cement_strength_scatter.png"), plot2_3, width = 9, height = 6, dpi = 300)
```

**Comment on results, paying attention to aesthetic mappings:** this
scatterplot maps four variables at once: `Cement` and
`Concrete_compressive_strength` are mapped to the x- and y-axes, `Age` is
mapped to point size, and `Water` is mapped to a colorblind-safe viridis
color scale, rather than encoding any of this meaning by color alone, shape
and position carry information too. The plot shows a general positive
relationship between cement content and strength, but the relationship is
far from perfectly linear; water content and age both visibly interact with
it. Mixtures with lower water content (darker purple points) tend to reach
higher strength values for a given amount of cement, which is consistent
with the well-known effect of the water-to-cement ratio on concrete
strength. Larger points (older samples) are also more common toward the
upper end of the strength range, reflecting the strength-gain-over-time
pattern seen in Section 2 above.

## Interactive Visualization

The static scatterplot above maps four variables (`Cement`,
`Concrete_compressive_strength`, `Age`, `Water`) at once, which makes it hard
to read exact values for any individual sample once points overlap. I
rebuild it with `plotly` so the reader can hover over any point to see its
exact cement content, strength, age, and water content. This is something a
static image cannot offer: with 1030 overlapping points, a static label for
every point would be illegible, while a hover-based interactive version lets
the reader inspect exactly the points they are curious about, on demand,
without cluttering the chart for everyone else.

```{r interactive-concrete, fig.width=9, fig.height=6}
p_interactive <- ggplot(new_concrete, aes(
  x = Cement,
  y = Concrete_compressive_strength,
  size = Age,
  color = Water,
  text = paste0(
    "Cement: ", Cement, " kg/m³",
    "<br>Strength: ", round(Concrete_compressive_strength, 1), " MPa",
    "<br>Age: ", Age, " days",
    "<br>Water: ", Water, " kg/m³"
  )
)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(option = "viridis") +
  labs(
    title = "Exploring Strength versus (Cement, Age, and Water) — interactive",
    x = "Cement",
    y = "Strength"
  ) +
  theme_minimal()

interactive_concrete_plot <- ggplotly(p_interactive, tooltip = "text")
interactive_concrete_plot
```

```{r save-interactive-concrete}
# Save a self-contained HTML version of the interactive chart, so it can be
# opened and explored independently of the knitted report.
htmlwidgets::saveWidget(interactive_concrete_plot, file = here::here("project-03", "concrete_interactive.html"), selfcontained = TRUE)
```

## Chart Redesign: Before / After

For the required redesign exercise, I take an uglier, earlier draft version
of the compressive strength histogram from Section 2.1 and fix it.

### Before: The Problematic Version

```{r redesign-before, fig.alt="An earlier, problematic version of the compressive strength histogram: 20 wide bins on a linear scale, filled entirely in a single flat red color with no colorblind-safe encoding, no axis units in the labels, and no title describing what the units on either axis represent."}
plot_redesign_before <- ggplot(new_concrete, aes(x = Concrete_compressive_strength)) +
  geom_histogram(bins = 20, fill = "red") +
  theme_minimal() +
  labs(title = "Bad Visualization (No transformation)")

plot_redesign_before
```

```{r save-redesign-before}
ggsave(here::here("figures", "concrete_redesign_before.png"), plot_redesign_before, width = 7, height = 5, dpi = 300)
```

**What is wrong with this chart:** the fill color is a flat, saturated red
with no analytical purpose, it does not encode any variable, and pure red is
one of the worst color choices for colorblind accessibility since it is
difficult to distinguish from dark backgrounds or other warm hues for some
forms of color vision deficiency. The title "Bad Visualization (No
transformation)" is also not an actual, informative title; it describes the
exercise rather than the data. Neither axis label specifies units (MPa,
number of samples), so a reader unfamiliar with the dataset cannot tell what
is being measured. With only 20 wide bins on a linear x-axis, the histogram
also loses some of the finer detail in the right-skewed tail of the
distribution.

### After: The Redesigned Version

```{r redesign-after, fig.alt="The redesigned histogram of concrete compressive strength: a colorblind-safe viridis fill, a log-scaled x-axis, finer 30 bins, fully labeled axes with units, and an informative title."}
plot_redesign_after <- ggplot(new_concrete, aes(x = Concrete_compressive_strength)) +
  geom_histogram(bins = 30, fill = viridis(1, option = "magma"), color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of Concrete Compressive Strength (Log Scale)",
    x = "Compressive strength (MPa, log scale)",
    y = "Number of samples"
  ) +
  theme_minimal()

plot_redesign_after
```

```{r save-redesign-after}
ggsave(here::here("figures", "concrete_redesign_after.png"), plot_redesign_after, width = 7, height = 5, dpi = 300)
```

**How the redesign fixes the problems:**

1. **Colorblind-safe fill**: the flat red is replaced with a single color
   drawn from the viridis/magma palette, which is designed to remain
   distinguishable under common forms of color vision deficiency, rather
   than an arbitrary, purely decorative red.
2. **Informative title and fully labeled, unit-bearing axes**: the title now
   names the variable being shown, and both axes specify their units (MPa
   for strength, a count of samples for the y-axis), so the chart is
   interpretable without needing the surrounding report text.
3. **Log-scaled x-axis with finer binning**: 30 bins on a log scale reveal
   more structure in the right-skewed tail of the distribution than 20
   wide, linearly-spaced bins did, without changing the underlying data.

## Conclusion

This project let me apply density plots, ridgeline plots, faceting, an
interactive `plotly` chart, and a deliberate before/after chart redesign to
two very different datasets: daily weather at Tampa International Airport
and concrete compressive strength testing. Across both datasets, the
recurring theme was that color, scale transformations (log scales in
particular), and explicit annotation (quantile lines, hover tooltips, units
in axis labels) consistently made the underlying patterns easier to read
without changing the data itself.
